Proof Modularization: Enabling Verified Design-Oriented Modifications in Cache Coherence Implementation

### 2014-01-30 paper/Introduction.tex
Abstract added
Clean up the intro in light of the abstract. Sounds a little more natural now.

Since the correctness of this abstraction is paramount in the correct
functioning of the system, significant effort is placed in making sure
that the cache coherence engine is correct. In practice, this is done
in two steps. First, a cache coherence protocol is constructed and
verified. This protocol generally directly represents many of
the key concurrent aspects of an implementation, \eg{} reads and
writes to the system are broken into local atomic micro-operations, but
ignores implementation details \eg{} buffer sizing and memory access
times. Second, once the protocol has been verified it is handed to
hardware experts who fill in the missing implementation details and
realize it at the Register Transfer Level (RTL) of gates, wires, and
registers. This involves further breaking down of local micro-operations to
make it realizable in hardware. This effort is not a one time translation, as
performance properties are not easily estimated; a significant amount of
performance testing is requires to select a good implementation.

From a methodological perspective, this two step process represent a
significant methodological failure. We spent a significant amount of
time verifying the protocol and then invalidate any correctness
guarantees to the implementation by performing not only a non-trivial
translation from the protocol representation to hardware, but also
the refinement to add missing implementation details.

Because verified protocols are often described via Term Rewrite
Systems (TRS).  it was shown by Dave \etal{} that a verified protocol
could be transliterated into Bluespec, a hardware description language
based on Guarded Atomic Actions, a restriction of
TRS~\cite{Hoe:TCAD,DNA:CoherenceImplementation}. This reduces the
correctness of the implementation to the correctness of the protocol,
effectively removing any implementation bugs. Though in the current
versions of the Bluespec language, each atomic action corresponds to
one hardware clock cycle, there has been some work to reduce this
restriction~\cite{Karczmarek}.

The issue with this protocol transliteration approach is that design
of a hardware system requires many design tweaks to reach optimal
performance as one cannot practically know the correct low-level
hardware results without evaluating the low-level representation.
Each of these changes can potentially breaks correctness and a full
reverification of correctness is needed in principle. As hardware
designers are unlikely to have the expertise or skill this effectively
nullifies the verification guarantee in design.

To fully realize this approach of design we need to provide not just a
single proof of correctness, but a proof framework covering any
variation of a design that might be of interest for performance
reasons. This would enable implementation level changes without
compromising on the ability to verify correctness of the design.

@@

In this paper, we describe an approach to provide this proof framework
for any particular cache coherence protocol via the example of a
MSI-style protocol and show how it can be extended to other protocols
like MESI and MOESI. We show how the modular structure of the proof
and design can be shared and how modification of microarchitectural
aspects of design directly translates to just reverification of our
local building block properties. We use Coq for our proof
framework. We also show how an atomic action based MSI protocol
description can be directly transliterated into an Inductive
definition in Coq, and verify its local building block properties
formally, thus completing the picture.

>>>>

Since the correctness of this abstraction is paramount in the correct
functioning of the system, significant effort is placed in making sure
that the cache coherence engine is correct. In practice, this is done
in two steps. First, a cache coherence protocol is constructed and
verified. This protocol generally directly represents many of
the key concurrent aspects of an implementation, \eg{} reads and
writes to the system are broken into local atomic micro-operations, but
ignores implementation details \eg{} buffer sizing and memory access
times. Second, once the protocol has been verified it is handed to
hardware experts who fill in the missing implementation details and
realize it at the Register Transfer Level (RTL) of gates, wires, and
registers. This involves further breaking down of local micro-operations to
make it realizable in hardware. This effort is not a one-time translation, as
performance properties are not easily estimated; a significant amount of
performance testing is requires to select a good implementation.

From a methodological perspective, this two-step process represent a
significant methodological failure. The significant effort of
verification is effectively lost by the concurrent non-trivial tasks
of translation from the protocol representation to a hardware
description and the necessary refinement to add missing implementation
details to the protocol.

Because verified protocols are often described via Term Rewrite
Systems (TRS). It was shown by Dave
\etal{}~\cite{DNA:CoherenceImplementation} that a verified protocol
could be transliterated into Bluespec, a hardware description language
based on Guarded Atomic Actions, a restriction of
TRS~\cite{Hoe:TCAD}. This reduces the correctness of the
implementation to the correctness of the protocol, effectively
removing any implementation bugs. While the current Bluespec compiler
has certain restrictions which limit the expressiveness of certain
operations \eg{} fairness is not guaranteed, atomic actions are always
implemented in a single cycle, work has been done to remove these
restrictions~\cite{DNA:CoherenceImplementation, Karczmarek}.

While better, this approach of protocol transliteration has
flaws. Design of a hardware system requires many tweaks to reach
optimal performance as one cannot practically know the correct
low-level hardware results without evaluating the low-level
representation.  Each of these changes can potentially breaks
correctness and a full reverification of correctness is needed in
principle. As hardware designers are unlikely to have the expertise or
skill to do make this verification, this greatly weakens the ability
to carry over our verification guarantee to the design.

For this approach to be truly viable, we need not a single proof of
correctness of a single design, but a proof framework covering any
variation of a design that might be of interest for performance
reasons. This would enable implementation level changes without
compromising on the ability to verify correctness of the design.

@@

In this paper, we describe an approach to provide this proof framework
for any particular cache coherence protocol via the example of a
MSI-style protocol and show how it can be extended to other protocols
like MESI and MOESI. We show how the modular structure of the proof
and design can be shared and how modification of microarchitectural
aspects of design directly translates to just reverification of our
local building block properties. We use Coq for our proof
framework. We also show how an atomic-action-based MSI protocol
description can be directly transliterated into an Inductive
definition in Coq, and verify its local building block properties
formally.

### 2014-01-30 paper/Title.tex
Abstract added
Clean up the intro in light of the abstract. Sounds a little more natural now.

\title{Stuff
}

@@

\author{
    Muralidaran Vijayaraghavan\\Massachusetts Institute of Technology\\vmurali@csail.mit.edu\and
    Nirav Dave\\SRI International\\ndave@csl.sri.com\and
    Arvind\\Massachusetts Institute of Technology\\arvind@csail.mit.edu
}

>>>>

\title{Proof Modularization: Enabling Verified Design-Oriented Modifications in Cache Coherence Implementation}

@@

\author{
%    Muralidaran Vijayaraghavan\\Massachusetts Institute of Technology\\vmurali@csail.mit.edu\and
%    Nirav Dave\\SRI International\\ndave@csl.sri.com\and
%    Adam Chlipala\\Massachusetts Institute of Technology\\adamc@csail.mit.edu\and
%    Arvind\\Massachusetts Institute of Technology\\arvind@csail.mit.edu
}
